\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-10-19 doc/paper-jss/sdmTMB-paper.Rnw
Abstract

\Abstract{
Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \INLA\ and estimation is conducted via maximum marginal likelihood with Template Model Builder (\TMB) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}

>>>>

\Abstract{
Geostatistical data are common across scientific fields and the dimensions of these data continue to become larger in space and time.
However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging for many users to implement.
Here, we introduce the \R\ package \sdmTMB, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, and \pkg{mgcv} to include spatial and spatiotemporal latent Gaussian Markov random fields using an SPDE-(stochastic partial differential equation) based approach.
SPDE matrices are constructed with \INLA\ and estimation is conducted via maximum marginal likelihood with Template Model Builder (\TMB) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}.
We describe the model and explore case studies that illustrate \sdmTMB's flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation).
Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation.
Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \sdmTMB\ can be an order of magnitude faster than \INLA.
We hope \sdmTMB's accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
}
